---
title: "Density Regressions - API-313M"
output: html_notebook
---

```{r, setwd, include=FALSE}
# Set working directory appropriately
working_dir <- "/Users/shg309/Dropbox (Personal)/Education/hks_cid_growth_lab/misc/gl_complexity_review"
knitr::opts_knit$set(root.dir = working_dir)
```

```{r, setup, include=FALSE}
# Load packages
packages <- 
    c(
        "tidyverse",
        "dataverse",
        "wbstats",
        "Matrix",
        "RCurl",
        "economiccomplexity",
        "modelsummary",
        "here"
    )
sapply(packages, library, character.only = T)
# setwd(working_dir)
here::i_am("proj/api_313m/linear_algebra.R")
source(here("proj", "api_313m", "setup.R"), local = knitr::knit_global())
```

This review session will cover the following topics:

- Recap of density and proximity
- Density and growth regressions
- Replication: Hausmann, R., Stock, D. P., & Yildrim, M. A. (2021). Implied comparative advantage. Research Policy, 104143.

# Recap of density and proximity

Proximity capture the relatedness of two entities, as in lecture 5:

Using minimum conditional probability method for calculating proximity
$$\phi_{pp'} = \frac{\sum_c M_{cp} M_{cp'}}{max(\sum_c{M_{cp}}, \sum_c{M_{cp'}})}$$

In matrix notation:

$$ \Phi = \min{(U^{-1} M M^T, M M^T U^{-1})} - \mathbb{1}_p $$

Where $U$ is the ubiquity matrix (ubiquity on diagonal), and $I_P$ is the identity matrix


Appears in several domains:

- International trade (proximities between products / countries)
- Cities
- Occupations / labor flows
- Patents
- Research papers

Density captures the overlap between the capabilities present in a country and those required by a product. It is the average proximity between the products that a country already makes ($M_{cp} > 1$), and the target product $p$.

$$d_{cp} = \frac{\sum_p M_{cp'}\phi_{pp'}}{\sum_p \phi_{pp'}}$$

In matrix notation:

$$ \delta = diag (\Phi \mathbb{1}_p)  $$
$$ D = M (\delta^{-1} M^T) $$

# Density regressions

The following snippet downloads trade data. If you're unable to download the file programmatically,
just go to [this link](https://www.dropbox.com/s/34a0tnboycvsnx0/country_hsproduct4digit_year.parquet) and download the data manually and read it in using `read_parquet`.

```{r}
# Download trade data
tradepath <-
  "https://intl-atlas-downloads.s3.amazonaws.com/country_hsproduct4digit_year.csv.zip"
proxpath <-
  "http://intl-atlas-downloads.s3.amazonaws.com/atlas_2_16_6/hs92_proximities.csv"
prodpath <-
  "https://intl-atlas-downloads.s3.amazonaws.com/hs_product.zip"
countrypath <-
  "https://intl-atlas-downloads.s3.amazonaws.com/location.zip"
# Set up folders
check_and_create_dir <- function(dirpath) {
  if (!dir.exists(dirpath)) {
    dir.create(dirpath)
  }
}
check_and_create_dir(here("data/raw"))
check_and_create_dir(here("data/processed"))

# Filepaths
trade_local_path <-
  here("data/raw/country_hsproduct4digit_year.csv.zip")
prox_local_path <- here("data/raw/hs92_proximities.csv")
prod_local_path <- here("data/raw/hs_product.zip")
country_local_path <- here("data/raw/location.zip")
if (!file.exists(trade_local_path)) {
  # Download
  download.file(
    url = tradepath,
    destfile = trade_local_path,
    method = "libcurl",
    timeout = 1200
  )
  download.file(url = prodpath,
                destfile = prod_local_path,)
  download.file(url = countrypath,
                destfile = country_local_path,)
  # Once downloaded, save it locally so you don't have to download each time
  trade <-
    read_csv(trade_local_path) %>%
    select(location_code, hs_product_code, year, export_value) %>%
    write_csv(here("data/processed/exports_data.csv"))
  prod <- read_csv(prod_local_path) %>%
    select(-product_id) %>%
    write_csv(here("data/processed/hs_prod.csv"))
  country <- read_csv(country_local_path) %>%
    select(-location_id) %>%
    write_csv(here("data/processed/country.csv"))
}
if (!file.exists(prox_local_path)) {
  download.file(
    url = proxpath,
    destfile = prox_local_path,
    method = "libcurl",
    timeout = 1200
  )
  prox <-
    read_csv(here("data/raw/hs92_proximities.csv")) %>%
    write_csv(here("data/processed/prox.csv"))
}

```

Now that we have the trade data downloaded, we will now recreate the metrics used in the paper: Hausmann, R., Stock, D. P., & Yildrim, M. A. (2021). [Implied comparative advantage](https://www.sciencedirect.com/science/article/pii/S0048733320302183). Research Policy, 104143.



```{r}
base_year <- 2010
end_year <- 2020

# Read in trade data and remove services
prox <- read_csv(here("data/processed/prox.csv"))
prod <- read_csv(here("data/processed/hs_prod.csv")) %>%
  rename(product = hs_product_code, product_name = hs_product_name_short_en)
country <- read_csv(here("data/processed/country.csv")) %>%
  rename(country = location_code, country_name = location_name_short_en)

trade <- read_csv(here("data/processed/exports_data.csv")) %>%
  filter(year >= base_year,
         hs_product_code %in% unique(prox$commoditycode_1))

# Rectangularize data
trade <- trade %>%
  tidyr::complete(year, location_code, hs_product_code, fill = list(export_value =
                                                                      0)) %>%
  rename(country = location_code,
         product = hs_product_code,
         value = export_value)

# Compute complexity metrics
bi <- balassa_index(trade, discrete = FALSE)
rca <- bi %>%
  as.matrix() %>%
  as_tibble() %>%
  mutate(country = rownames(bi)) %>%
  pivot_longer(-country, names_to = "product", values_to = "rca")

# Get ECI, PCI
cm <- complexity_measures(bi, method = 'reflections')
eci <- enframe(cm$complexity_index_country) %>%
  rename(country = name, eci = value) %>%
  left_join(country)
pci <- enframe(cm$complexity_index_product) %>%
  rename(product = name, pci = value) %>%
  left_join(prod)

# Get WDI data
# Population, GDP per capita, GDP
wb_indicators <- c(gdp_capita = "NY.GDP.PCAP.CD",
                   pop = "SP.POP.TOTL",
                   gdp = "NY.GDP.MKTP.CD")

wdi <-
  wb_data(wb_indicators, start_date = base_year, end_date = end_year + 1) %>%
  select(-iso2c, -country) %>%
  rename(country = iso3c, year = date) %>%
  filter(!is.null(pop)) %>%
  filter(year >= base_year, year <= end_year)

# Add these back to the trade dataset
trade_cm <- trade %>%
  left_join(rca) %>%
  left_join(select(eci, -country_name)) %>%
  left_join(select(pci, -product_name)) %>%
  inner_join(wdi)

# Get rpca
trade_cm <- trade_cm %>%
  group_by(year, product) %>%
  mutate(prod_total = sum(value, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(pop_total = sum(pop, na.rm = TRUE)) %>%
  mutate(rpca = (value / pop) / (prod_total / pop_total)) %>%
  select(-prod_total, -pop_total)
# mutate(rpca = ifelse(rpca >= 1, 1, 0))

# Repeat for each year
selected_year <- 2010

# Convert rpca to matrix
rpca_mat <- trade_cm %>%
  filter(year == selected_year) %>%
  mutate(rpca = replace_na(rpca, 0)) %>%
  pivot_wider(
    id_cols = country,
    names_from = product,
    values_from = rpca,
    values_fill = 0
  )
country_list <- rpca_mat$country
rpca_mat <- rpca_mat %>%
  select(-country)
prod_list <- colnames(rpca_mat)
rpca_mat <- as.matrix(rpca_mat)
colnames(rpca_mat) <- prod_list
rownames(rpca_mat) <- country_list

# Check if any rows or cols sum to zero and remove
rows_to_remove <- enframe(rowSums(rpca_mat)) %>%
  filter(value == 0) %>%
  pull(name)
cols_to_remove <- enframe(colSums(rpca_mat)) %>%
  filter(value == 0) %>%
  pull(name)
rpca_mat <-
  rpca_mat[!rownames(rpca_mat) %in% rows_to_remove, !colnames(rpca_mat) %in% cols_to_remove]
prod_list <- colnames(rpca_mat)
country_list <- rownames(rpca_mat)

# Mean rpca by location
location_mean_rpca <- trade_cm %>%
  group_by(year, country) %>%
  summarise(location_mean_rpca = mean(rpca, na.rm = TRUE))


# Compute proximities as per paper
prox_product <- (1 + cor(rpca_mat)) * .5
rownames(prox_product) <- prod_list
colnames(prox_product) <- prod_list

prox_country <- (1 + cor(t(rpca_mat))) * .5
rownames(prox_country) <- country_list
colnames(prox_country) <- country_list

# Compute densities as per paper
prox_product_sum <- colSums(prox_product)
density_prod <- rpca_mat %*% (t(prox_product) / prox_product_sum)

prox_country_sum <- colSums(prox_country)
density_country <-
  t(rpca_mat) %*% (t(prox_country) / prox_country_sum)

density_prod <- density_prod %>%
  as.data.frame()
density_country <- density_country %>%
  as.data.frame()

density_prod <- density_prod %>%
  mutate(country = rownames(density_prod)) %>%
  pivot_longer(-country, names_to = "product", values_to = "density_prod")

density_country <- density_country %>%
  mutate(product = rownames(density_country)) %>%
  pivot_longer(-product, names_to = "country", values_to = "density_country")

# Compute growth rates
exports <- trade_cm %>%
  select(year, country, product, value) %>%
  filter(year %in% c(2010, 2020)) %>%
  pivot_wider(
    id_cols = c(country, product),
    names_from = year,
    values_from = value
  ) %>%
  filter(`2010` > 0, `2020` > 0) %>%
  mutate(growth_rate = log(`2020` / `2010`) / (2020 - 2010))

product_sum <- exports %>%
  select(-country) %>%
  group_by(product) %>%
  mutate_all(~ sum(.x, na.rm = TRUE))
country_sum <- exports %>%
  select(-product) %>%
  group_by(country) %>%
  mutate_all(~ sum(.x, na.rm = TRUE))
exports <- exports %>%
  mutate(radial_product_growth = log(product_sum$`2020` / product_sum$`2010`) / (2020 -
                                                                                   2010)) %>%
  mutate(radial_country_growth = log(country_sum$`2020` / country_sum$`2010`) / (2020 -
                                                                                   2010)) %>%
  mutate(
    base_year_country_total = country_sum$`2010`,
    base_year_product_total = product_sum$`2010`,
    base_year_exports = `2010`
  ) %>%
  select(-`2010`, -`2020`)

# Merge everything back
trade_cm <- trade_cm %>%
  inner_join(exports) %>%
  inner_join(location_mean_rpca) %>%
  inner_join(density_prod) %>%
  inner_join(density_country)

# Remove infs
trade_cm <- trade_cm %>%
  filter(is.finite(log(growth_rate))) %>%
  filter(is.finite(log(rpca)))

```

Notice that we calculated a (perhaps) new metric called `rpca`, or `RpCA`, which is the population-weighted market share. In some cases, rpca acts as a better metric than RCA as it weights by population instead of product market size. However, large countries such as India and China are disadvantaged by the RpCA metric.

The densities (country-space density and product-space density) are then used to get a sense of how far a country-product is from its "expected" level of trade. The residuals from the first stage regressions are used to predict future growth in exports in the country-product.


```{r}
# Run regression
reg_hybrid <-
    lm(
        log(rpca) ~ log(density_prod) + log(density_country),
        data = trade_cm,
        na.action = na.exclude
    )
reg_prod <-
    lm(log(rpca) ~ log(density_prod),
        data = trade_cm,
        na.action = na.exclude
    )
reg_country <-
    lm(log(rpca) ~ log(density_country),
        data = trade_cm,
        na.action = na.exclude
    )

first_stage_models <- list(
    "(1)" = reg_prod,
    "(2)" = reg_country,
    "(3)" = reg_hybrid
)

modelsummary(
    first_stage_models,
    fmt = 2,
    estimate  = "{estimate} ({std.error}){stars}"
)

modelsummary(
    first_stage_models,
    fmt = 2,
    estimate  = "{estimate} ({std.error}){stars}",
    output = here("proj/api_313m/tables/first_stage.txt")
)

# Second stage
trade_cm <- trade_cm %>%
    mutate(
        residual_hybrid_density = resid(reg_hybrid),
        residual_prod_space_density = resid(reg_prod),
        residual_country_space_density = resid(reg_country)
    )

second_stage_models <- list(
    "(1)" = lm(
        log(growth_rate) ~ residual_prod_space_density,
        data = trade_cm,
        na.action = na.exclude
    ),
    "(2)" = lm(
        log(growth_rate) ~ residual_country_space_density,
        data = trade_cm,
        na.action = na.exclude
    ),
    "(3)" = lm(
        log(growth_rate) ~ residual_hybrid_density,
        data = trade_cm,
        na.action = na.exclude
    ),
    "(4)" = lm(
        log(growth_rate) ~ residual_hybrid_density + log(base_year_exports) + log(pop) + log(base_year_product_total) + location_mean_rpca,
        data = trade_cm,
        na.action = na.exclude
    ),
    "(5)" = lm(
        log(growth_rate) ~ log(base_year_exports) + log(pop) + log(base_year_product_total) + location_mean_rpca + radial_product_growth + radial_country_growth,
        data = trade_cm,
        na.action = na.exclude
    ),
    "(6)" = lm(
        log(growth_rate) ~ residual_hybrid_density + log(base_year_exports) + log(pop) + log(base_year_product_total) + location_mean_rpca + radial_product_growth + radial_country_growth,
        data = trade_cm,
        na.action = na.exclude
    )
)

modelsummary(
    second_stage_models,
    fmt = 2,
    estimate  = "{estimate} ({std.error}){stars}"
)

modelsummary(
    second_stage_models,
    fmt = 2,
    estimate  = "{estimate} ({std.error}){stars}",
    output = here("proj/api_313m/tables/second_stage.txt")
)
```